OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-03-22 paper/paper.Rmd
Edit algorithm description w/ feedback from SLW (@mothur-westcott)

- **TODO: ask Sarah Westcott to check the accuracy of this description**.

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
their sequences, a list of query sequences to assign to the reference OTUs, and
the sequence pairs that are within the distance threshold (e.g. 0.03).
Query sequences are randomly seeded in reference OTUs, then for each sequence
the algorithm calculates the quality metric (default: MCC) based on whether the
sequence stays in its current OTU, moves to each of the other OTUs, or is
discarded.
If two or more OTU assignments are of equal quality, a random number generator
is used to break the tie.
This process is repeated until the quality metric stabilizes, changing by no
more than 0.0001 by default, or until a maximum number of iterations is reached
(default: 100).
In closed-reference mode, any query sequences that cannot be assigned are thrown
out (**TODO: exactly what determines whether a seq can't be assigned?**), and
the results only contain OTUs that exist in the original reference.
In open-reference mode, unassigned query sequences are clustered _de novo_ using
OptiClust to generate additional OTUs. The final quality score is reported with
the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
reference OTUs of an independent database, or 2) split the dataset into a
reference and query fraction, cluster the reference sequences _de novo_, then
fit the query sequences to the reference OTUs.

@@

- TODO: start paragraph with question, end with why we should care. transitions to move the story along.
- start:"In order to compare Optifit.."
- end: "optifit performs better"
- AND, BUT, THEREFORE, structure
- order of importance: mcc, mapping, sec

**TODO: lead with comparison of optifit vs current state of the art. give context at beginning.**
We performed _de novo_ clustering with OptiClust at distance threshold of 3% (TODO R CODE HERE)
using the Greengenes database, SILVA non-redundant database, or the Ribosomal
Database Project (RDP) [@desantis_greengenes_2006; @quast_silva_2013;
@cole_ribosomal_2014].
We then performed refernce clustering with OptiFit, fitting each dataset to each 
set of database OTUs repeated 100 times.
We measured the algorithm run time and evaluated OTU quality using the MCC as
described previously [@westcott_opticlust_2017].
To compare to existing software, we also clustered each dataset into _de novo_
OTUs  using OptiClust and VSEARCH, and with VSEARCH in reference-based mode
against the Greengenes database.

>>>>

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
their sequences, a list of query sequences to assign to the reference OTUs, and
the sequence pairs that are within the distance threshold (e.g. 0.03).
Initially, query sequences are placed in singleton OTUs. 
Then, the algorithm iteratively reassigns the query sequences to the reference 
OTUs to optimize the Matthews correlation coefficient (MCC). 
A sequence will remain unassigned if the MCC value is maximized when the
sequence is a singleton rather than assigned to a reference OTU.
This process is repeated until the MCC stabilizes, changing by no more than
0.0001 by default, or until a maximum number of iterations is reached (default:
100).
In closed-reference mode, any query sequences that cannot be assigned to
references OTUs are discarded, and the results will only contain OTUs that exist
in the original reference.
In open-reference mode, unassigned query sequences are clustered _de novo_ using
OptiClust to generate additional OTUs. 
The final quality score is reported with the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
reference OTUs of an independent database, or 2) split the dataset into a
reference and query fraction, cluster the reference sequences _de novo_, then
fit the query sequences to the reference OTUs.

@@

- TODO: start paragraph with question, end with why we should care. transitions to move the story along.
- start:"In order to compare Optifit.."
- end: "optifit performs better"
- AND, BUT, THEREFORE, structure
- order of importance: mcc, mapping, sec (1 paragraph each, in this order)

**TODO: lead with comparison of optifit vs current state of the art. give context at beginning.**
We performed _de novo_ clustering with OptiClust at distance threshold of 3%
using the Greengenes database, SILVA non-redundant database, or the Ribosomal
Database Project (RDP) [@desantis_greengenes_2006; @quast_silva_2013;
@cole_ribosomal_2014].
We then performed refernce clustering with OptiFit, fitting each dataset to each 
set of database OTUs repeated 100 times.
We measured the algorithm run time and evaluated OTU quality using the MCC as
described previously [@westcott_opticlust_2017].
To compare to existing software, we also clustered each dataset into _de novo_
OTUs  using OptiClust and VSEARCH, and with VSEARCH in reference-based mode
against the Greengenes database.
